The role of learning in the evolution of status signaling: a modeling approach

### 2023-11-24 manuscript/manuscript_2.0.Rmd
starting revision changes

The decision of whether to escalate a fight against an interacting 
partner can depend on the size of the partner's badge, 
with the dependency being determined by the focal's experiences acquired
through a learning process. We implement learning using 
the actor-critic approach from reinforcement learning (RL) theory [@sutton_Reinforcement_2018;@quinones_Reinforcement_2020; @leimar_Learning_2019;
@leimar_Evolution_2021].
Individuals estimate the reward (pay-off) expected from  interacting 
with partners of different badge sizes (the critic in RL terminology). 
After each interaction they update the estimate of reward proportionally 
to the difference between their current estimate and the observed 
reward (prediction error $\delta$) and to the speed of learning 
($\mathrm{A}$). Furthermore, individuals express different probabilities of 
retreating/attacking depending on the badge size of their opponent 
(the actor in RL). They update the probability of 
retreating/attacking depending on whether retreating/attacking
leads to an increase in the reward estimation. 
Thus, if a focal individual decides to escalate a fight against 
an individual with a small badge and this leads to an increase in 
the reward estimation, then the focal individual will increase the probability 
of escalating fights with individuals of small badges in the future. 
Given that badge size is a real number between 0 and 1, there are infinitely 
many badge sizes. Thus, the reward estimation, as well as the probability 
of retreating/attacking, must be generalized across different values. 
To implement generalization we use the linear function approximation 
method based on radial basis functions [@sutton_Reinforcement_2018].
Specifically, we pick $c$ feature centres, which are evenly spaced values 
along the badge size interval ([0,1]) where the updates are focused. 
For simplicity, we keep the location of these feature centres 
constant and  the same for all individuals; they are stored in 
vector $\symbf{b}$ . Each of these feature centres 
is associated with a weight for reward estimation and tendency to play 
retreat in a given interaction. The reward estimation and the tendency 
to retreat are calculated (in every interaction) as the sum of the weights 
associated with each feature centre, weighted by the response 
triggered by the feature (Fig. \ref{fig:modelStruc} B, dots 
represent the feature weights). The response of each feature centre 
diminishes as a Gaussian function with the distance between the 
feature centre and the badge size of 
the partner (Fig. \ref{fig:modelStruc} B, grey line). Formally, the 
reward estimation $\hat{R}$ when the focal individual ($i$) faces 
individual $j$ is given by,
\
\begin{equation}
\hat{R}_{ij} =  \sum_{z=1}^{c} x_z e^{(-\frac{|B_i-b_z|}{2\theta^2})} 
\end{equation}

@@

Individuals interact as the focal individual $n$ times and the interaction 
partner is chosen at random. Thus, the expected number of interactions for 
each individual is $n+\frac{n(N+1)}{N}$. 
After the interaction round , individuals in the population 
reproduce with a probability proportional to their total pay-off $w_i$, which is a sum of the 
baseline pay-off ($w_0$) and all the pay-offs obtained throughout their 
life. Thus, the combination of natural selection and genetic drift 
changes the distribution of values in $\alpha$ and $\beta$ that segregates 
in the population, effectively changing  the badges expressed and the 
communication system.

@@

```{r handicap,echo=FALSE,out.width='100%',fig.height=6,fig.align='center',fig.show='hold',fig.cap="Evolution of the badge size as a handicap mediated by learning. Top panels (A.1 and A.2) show the evolutionary dynamics of the sender code. On the left (A.1), changes in the distribution of values for the intercept of the reaction norm ($\\alpha$); on the right (A.2) changes in the distribution of the slope ($\\beta$). Darker areas of the background correspond to values with high frequency, while orange lines in both panels show the mean of the distribution. X axis in the evolutionary dynamics are given in thousands of generations. Grey lines show the generation time corresponding to the panels below portraing the sender (B) and receiver code(C). In the bottom panels (C1-4), the learned reaction norms correspond to individuals in the population after the interaction round. Color scale indicates quality just as in Fig. \\ref{fig:modelStruc}. On the first half of the simulation (B.1 and B.2) the badge size evolves to its minimum value. In the second half, the slope takes positive values and reaction norms evolve to produce an honest signal of quality. Correspondingly, individuals learn to react to badge size by increasing the probability of retreating with increasing badge sizes"  ,warning=FALSE}

@@

axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

```{r branching,echo=FALSE,out.width='100%',fig.height=6,fig.align='center',fig.show='hold',fig.cap= "The evolution of cheap signals. Portrait of the evolutionary dynamics of the sender code with snapsots of both sender and receiver codes just as in fig. \\ref{fig:handicap}. The top panels (A) show changes in the distribution of values for $\\alpha$ (A.1) and $\\beta$ (A.2) along evolutionary time. Panels below correspond to snapshots of the sender (B) and receiver codes (C). Generation time of the snapshots are indicated by the grey lines in the top panels. The unimodal distribution with which $\\alpha$ (A.1) starts the simulation, quickly turns into a bimondal distribution, and at about 10000 generations it splits further into three modes. These three peaks correspond to three type of reaction norms in panels B.1-3."}

@@

axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

>>>>

The decision of whether to escalate a fight against an interacting 
partner can depend on the size of the partner's badge, 
with the dependency being determined by the focal's experiences acquired
through a learning process. We implement learning using 
the actor-critic approach from reinforcement learning (RL) theory [@sutton_Reinforcement_2018;@quinones_Reinforcement_2020; @leimar_Learning_2019;
@leimar_Evolution_2021].
Individuals estimate the reward (pay-off) expected from  interacting 
with partners of different badge sizes (the critic in RL terminology). 
After each interaction they update the estimate of reward proportionally 
to the difference between their current estimate and the observed 
reward (prediction error $\delta$) and to the speed of learning 
($\mathrm{A}$). Furthermore, individuals express different probabilities of 
retreating/attacking depending on the badge size of their opponent 
(the actor in RL). They update the probability of 
retreating/attacking depending on whether retreating/attacking
leads to an increase in the reward estimation. 
Thus, if a focal individual decides to escalate a fight against 
an individual with a small badge and this leads to an increase in 
the reward estimation, then the focal individual will increase the probability 
of escalating fights with individuals of small badges in the future. 
Given that badge size is a real number between 0 and 1, there are infinitely 
many badge sizes. Thus, the reward estimation, as well as the probability 
of retreating/attacking, must be generalized across different values. 
To implement generalization we use the linear function approximation 
method based on radial basis functions [@sutton_Reinforcement_2018].
Specifically, we pick $c$ feature centres, which are evenly spaced values 
along the badge size interval ([0,1]) where the updates are focused. 
For simplicity, we keep the location of these feature centres 
constant and  the same for all individuals; they are stored in 
vector $\symbf{b}$ . Each of these feature centres 
is associated with a weight for reward estimation and tendency to play 
retreat in a given interaction. The reward estimation and the tendency 
to retreat are calculated (in every interaction) as the sum of the weights 
associated with each feature centre, weighted by the response 
triggered by the feature (Fig. \ref{fig:modelStruc} B, dots 
represent the feature weights). The response of each feature centre 
diminishes as a Gaussian function with the distance between the 
feature centre and the badge size of 
the partner (Fig. \ref{fig:modelStruc} B, grey line). Formally, the 
reward estimation $\hat{R}$ when the focal individual ($i$) faces 
individual $j$ is given by,
\
\begin{equation}
\hat{R}_{ij} =  \sum_{z=1}^{c} x_z e^{(-\frac{|B_j-b_z|}{2\theta^2})} 
\end{equation}

@@

Individuals interact as the focal individual $n$ times and the interaction 
partner is chosen at random. Thus, the expected number of interactions for 
each individual is $2n$. 
After the interaction round , individuals in the population 
reproduce with a probability proportional to their total pay-off $w_i$, which is a sum of the 
baseline pay-off ($w_0$) and all the pay-offs obtained throughout their 
life. Thus, the combination of natural selection and genetic drift 
changes the distribution of values in $\alpha$ and $\beta$ that segregates 
in the population, effectively changing  the badges expressed and the 
communication system.

@@

```{r handicap,echo=FALSE,out.width='100%',fig.height=7,fig.align='center',fig.show='hold',fig.cap="Evolution of the badge size as a handicap mediated by learning. Top panels (A.1 and A.2) show the evolutionary dynamics of the sender code. In the y axis the trait value, in the x axis the generation time in thousands of generations. On the left (A.1), changes in the distribution of values for the intercept of the reaction norm ($\\alpha$); on the right (A.2) changes in the distribution of the slope ($\\beta$). Darker areas of the background correspond to values with high frequency, while orange lines in both panels show the mean of the distribution. Grey lines show the generation time corresponding to the panels below portraing the sender (B) and receiver code (C). In the bottom panels (C1-4), the learned reaction norms correspond to individuals in the population after the interaction round. Color scale indicates quality just as in Fig. \\ref{fig:modelStruc}. The black dashed line shows the average reaction norm of the population. On the first half of the simulation (B.1 and B.2) the badge size evolves to its minimum value. In the second half, the slope takes positive values and reaction norms evolve to produce an honest signal of quality. Correspondingly, individuals learn to react to badge size by increasing the probability of retreating with increasing badge sizes"  ,warning=FALSE}

@@

text(x = 2000,y=0,"time:")
axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

text(x = 2000,y=-3.1,"time:")
axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

```{r branching,echo=FALSE,out.width='100%',fig.height=6,fig.align='center',fig.show='hold',fig.cap= "The evolution of cheap signals. Portrait of the evolutionary dynamics of the sender code with snapsots of both sender and receiver codes just as in fig. \\ref{fig:handicap}. The top panels (A) show changes in the distribution of values for $\\alpha$ (A.1) and $\\beta$ (A.2) along evolutionary time. In the y axis the trait value, in the x axis the generation time in thousands of generations. Panels below correspond to snapshots of the sender (B) and receiver codes (C). Generation time of the snapshots are indicated by the grey lines in the top panels. Color of the lines indicates the quality of the individual, as indicated in the color code in C.1. The dashed line represents the average reaction norm of the popúlations. The unimodal distribution with which $\\alpha$ (A.1) starts the simulation, quickly turns into a bimondal distribution, and at about 10000 generations it splits further into three modes. These three peaks correspond to three type of reaction norms in panels B.1-3."}

@@

text(x = 2000,y=-4,"time:")
axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

text(x = 2000,y=-3.2,"time:")
axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

text(x = 2000,y=-1,"time:")
axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

text(x = 2000,y=-2,"time:")
axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

text(x = 2000,y=-0.4,"time:")
axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

@@

text(x = 2000,y=-1.5,"time:")
axis(side=1,padj = -2,at = axTicks(1)[2:(length(axTicks(1)))],
     labels = axTicks(1)[2:(length(axTicks(1)))]/1000,cex.axis=cexAxis,
     hadj = 1.3,line = -0.5,tick = FALSE)

### 2023-11-24 manuscript/manuscript_2.0.tex
starting revision changes

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
  \AddToHook{env/abstract/before}{\vspace*{-4.5cm}}
  \usepackage{float}
  \usepackage{graphicx}
  \usepackage{unicode-math}
  \usepackage{setspace}\doublespacing
  \usepackage[backref=page]{hyperref}
  \linespread{2}
  \usepackage{lineno}
  \linenumbers
  \floatplacement{figure}{H}
  \newcommand{\beginsupplement}{ \setcounter{table}{0}            
  \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={ The role of learning in the evolution of status signalling: a modeling approach},
  pdfkeywords={My keywords},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{\large \textbf{Title:} The role of learning in the evolution of
status signalling: a modeling approach}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\normalsize \textbf{Abbreviated title:} Learning and evolution
in status signalling}
\date{}

@@

@@

The decision of whether to escalate a fight against an interacting
partner can depend on the size of the partner's badge, with the
dependency being determined by the focal's experiences acquired through
a learning process. We implement learning using the actor-critic
approach from reinforcement learning (RL) theory (Sutton and Barto 2018;
Quiñones et al. 2020; Leimar and McNamara 2019; Leimar 2021).
Individuals estimate the reward (pay-off) expected from interacting with
partners of different badge sizes (the critic in RL terminology). After
each interaction they update the estimate of reward proportionally to
the difference between their current estimate and the observed reward
(prediction error \(\delta\)) and to the speed of learning
(\(\mathrm{A}\)). Furthermore, individuals express different
probabilities of retreating/attacking depending on the badge size of
their opponent (the actor in RL). They update the probability of
retreating/attacking depending on whether retreating/attacking leads to
an increase in the reward estimation. Thus, if a focal individual
decides to escalate a fight against an individual with a small badge and
this leads to an increase in the reward estimation, then the focal
individual will increase the probability of escalating fights with
individuals of small badges in the future. Given that badge size is a
real number between 0 and 1, there are infinitely many badge sizes.
Thus, the reward estimation, as well as the probability of
retreating/attacking, must be generalized across different values. To
implement generalization we use the linear function approximation method
based on radial basis functions (Sutton and Barto 2018). Specifically,
we pick \(c\) feature centres, which are evenly spaced values along the
badge size interval ({[}0,1{]}) where the updates are focused. For
simplicity, we keep the location of these feature centres constant and
the same for all individuals; they are stored in vector \(\symbf{b}\) .
Each of these feature centres is associated with a weight for reward
estimation and tendency to play retreat in a given interaction. The
reward estimation and the tendency to retreat are calculated (in every
interaction) as the sum of the weights associated with each feature
centre, weighted by the response triggered by the feature (Fig.
\ref{fig:modelStruc} B, dots represent the feature weights). The
response of each feature centre diminishes as a Gaussian function with
the distance between the feature centre and the badge size of the
partner (Fig. \ref{fig:modelStruc} B, grey line). Formally, the reward
estimation \(\hat{R}\) when the focal individual (\(i\)) faces
individual \(j\) is given by,\\
\begin{equation}
\hat{R}_{ij} =  \sum_{z=1}^{c} x_z e^{(-\frac{|B_i-b_z|}{2\theta^2})} 
\end{equation}

@@

Individuals interact as the focal individual \(n\) times and the
interaction partner is chosen at random. Thus, the expected number of
interactions for each individual is \(n+\frac{n(N+1)}{N}\). After the
interaction round , individuals in the population reproduce with a
probability proportional to their total pay-off \(w_i\), which is a sum
of the baseline pay-off (\(w_0\)) and all the pay-offs obtained
throughout their life. Thus, the combination of natural selection and
genetic drift changes the distribution of values in \(\alpha\) and
\(\beta\) that segregates in the population, effectively changing the
badges expressed and the communication system.

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_1.0_files/figure-latex/modelStruc-1}

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_1.0_files/figure-latex/handicap-1}

@@

\caption{Evolution of the badge size as a handicap mediated by learning. Top panels (A.1 and A.2) show the evolutionary dynamics of the sender code. On the left (A.1), changes in the distribution of values for the intercept of the reaction norm ($\alpha$); on the right (A.2) changes in the distribution of the slope ($\beta$). Darker areas of the background correspond to values with high frequency, while orange lines in both panels show the mean of the distribution. X axis in the evolutionary dynamics are given in thousands of generations. Grey lines show the generation time corresponding to the panels below portraing the sender (B) and receiver code(C). In the bottom panels (C1-4), the learned reaction norms correspond to individuals in the population after the interaction round. Color scale indicates quality just as in Fig. \ref{fig:modelStruc}. On the first half of the simulation (B.1 and B.2) the badge size evolves to its minimum value. In the second half, the slope takes positive values and reaction norms evolve to produce an honest signal of quality. Correspondingly, individuals learn to react to badge size by increasing the probability of retreating with increasing badge sizes}\label{fig:handicap}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_1.0_files/figure-latex/branching-1}

@@

\caption{The evolution of cheap signals. Portrait of the evolutionary dynamics of the sender code with snapsots of both sender and receiver codes just as in fig. \ref{fig:handicap}. The top panels (A) show changes in the distribution of values for $\alpha$ (A.1) and $\beta$ (A.2) along evolutionary time. Panels below correspond to snapshots of the sender (B) and receiver codes (C). Generation time of the snapshots are indicated by the grey lines in the top panels. The unimodal distribution with which $\alpha$ (A.1) starts the simulation, quickly turns into a bimondal distribution, and at about 10000 generations it splits further into three modes. These three peaks correspond to three type of reaction norms in panels B.1-3.}\label{fig:branching}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_1.0_files/figure-latex/honestDiv-1}

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_1.0_files/figure-latex/startCond-1}

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_1.0_files/figure-latex/noBadge-1}

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_1.0_files/figure-latex/learnEvol-1}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
  \AddToHook{env/abstract/before}{\vspace*{-4.5cm}}
  \usepackage{float}
  \usepackage{graphicx}
  \usepackage{unicode-math}
  \usepackage{setspace}\doublespacing
  \usepackage[backref=page]{hyperref}
  \linespread{2}
  \usepackage{lineno}
  \linenumbers
  \floatplacement{figure}{H}
  \newcommand{\beginsupplement}{ \setcounter{table}{0}            
  \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
  \providecommand{\keywords}[1]{\textbf{\textit{Keywords---}} #1}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The role of learning in the evolution of status signaling: a modeling approach},
  pdfkeywords={Biological signals, evolutionary branching, agent-based
modeling, animal communication, badges of status, handicap principle},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{The role of learning in the evolution of status signaling: a
modeling approach}
\date{}

@@

\keywords{Signals, evolutionary branching, agent-based modeling, animal communication, badges of status, handicap principle}

@@

The decision of whether to escalate a fight against an interacting
partner can depend on the size of the partner's badge, with the
dependency being determined by the focal's experiences acquired through
a learning process. We implement learning using the actor-critic
approach from reinforcement learning (RL) theory (Sutton and Barto 2018;
Quiñones et al. 2020; Leimar and McNamara 2019; Leimar 2021).
Individuals estimate the reward (pay-off) expected from interacting with
partners of different badge sizes (the critic in RL terminology). After
each interaction they update the estimate of reward proportionally to
the difference between their current estimate and the observed reward
(prediction error \(\delta\)) and to the speed of learning
(\(\mathrm{A}\)). Furthermore, individuals express different
probabilities of retreating/attacking depending on the badge size of
their opponent (the actor in RL). They update the probability of
retreating/attacking depending on whether retreating/attacking leads to
an increase in the reward estimation. Thus, if a focal individual
decides to escalate a fight against an individual with a small badge and
this leads to an increase in the reward estimation, then the focal
individual will increase the probability of escalating fights with
individuals of small badges in the future. Given that badge size is a
real number between 0 and 1, there are infinitely many badge sizes.
Thus, the reward estimation, as well as the probability of
retreating/attacking, must be generalized across different values. To
implement generalization we use the linear function approximation method
based on radial basis functions (Sutton and Barto 2018). Specifically,
we pick \(c\) feature centres, which are evenly spaced values along the
badge size interval ({[}0,1{]}) where the updates are focused. For
simplicity, we keep the location of these feature centres constant and
the same for all individuals; they are stored in vector \(\symbf{b}\) .
Each of these feature centres is associated with a weight for reward
estimation and tendency to play retreat in a given interaction. The
reward estimation and the tendency to retreat are calculated (in every
interaction) as the sum of the weights associated with each feature
centre, weighted by the response triggered by the feature (Fig.
\ref{fig:modelStruc} B, dots represent the feature weights). The
response of each feature centre diminishes as a Gaussian function with
the distance between the feature centre and the badge size of the
partner (Fig. \ref{fig:modelStruc} B, grey line). Formally, the reward
estimation \(\hat{R}\) when the focal individual (\(i\)) faces
individual \(j\) is given by,\\
\begin{equation}
\hat{R}_{ij} =  \sum_{z=1}^{c} x_z e^{(-\frac{|B_j-b_z|}{2\theta^2})} 
\end{equation}

@@

Individuals interact as the focal individual \(n\) times and the
interaction partner is chosen at random. Thus, the expected number of
interactions for each individual is \(2n\). After the interaction round
, individuals in the population reproduce with a probability
proportional to their total pay-off \(w_i\), which is a sum of the
baseline pay-off (\(w_0\)) and all the pay-offs obtained throughout
their life. Thus, the combination of natural selection and genetic drift
changes the distribution of values in \(\alpha\) and \(\beta\) that
segregates in the population, effectively changing the badges expressed
and the communication system.

The code necessary to run simulations as well as to produce
visualizations can be found in:
\url{https://doi.org/10.5281/zenodo.8359809}

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_2.0_files/figure-latex/modelStruc-1}

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_2.0_files/figure-latex/handicap-1}

@@

\caption{Evolution of the badge size as a handicap mediated by learning. Top panels (A.1 and A.2) show the evolutionary dynamics of the sender code. In the y axis the trait value, in the x axis the generation time in thousands of generations. On the left (A.1), changes in the distribution of values for the intercept of the reaction norm ($\alpha$); on the right (A.2) changes in the distribution of the slope ($\beta$). Darker areas of the background correspond to values with high frequency, while orange lines in both panels show the mean of the distribution. Grey lines show the generation time corresponding to the panels below portraing the sender (B) and receiver code (C). In the bottom panels (C1-4), the learned reaction norms correspond to individuals in the population after the interaction round. Color scale indicates quality just as in Fig. \ref{fig:modelStruc}. The black dashed line shows the average reaction norm of the population. On the first half of the simulation (B.1 and B.2) the badge size evolves to its minimum value. In the second half, the slope takes positive values and reaction norms evolve to produce an honest signal of quality. Correspondingly, individuals learn to react to badge size by increasing the probability of retreating with increasing badge sizes}\label{fig:handicap}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_2.0_files/figure-latex/branching-1}

@@

\caption{The evolution of cheap signals. Portrait of the evolutionary dynamics of the sender code with snapsots of both sender and receiver codes just as in fig. \ref{fig:handicap}. The top panels (A) show changes in the distribution of values for $\alpha$ (A.1) and $\beta$ (A.2) along evolutionary time. In the y axis the trait value, in the x axis the generation time in thousands of generations. Panels below correspond to snapshots of the sender (B) and receiver codes (C). Generation time of the snapshots are indicated by the grey lines in the top panels. Color of the lines indicates the quality of the individual, as indicated in the color code in C.1. The dashed line represents the average reaction norm of the popúlations. The unimodal distribution with which $\alpha$ (A.1) starts the simulation, quickly turns into a bimondal distribution, and at about 10000 generations it splits further into three modes. These three peaks correspond to three type of reaction norms in panels B.1-3.}\label{fig:branching}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_2.0_files/figure-latex/honestDiv-1}

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_2.0_files/figure-latex/startCond-1}

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_2.0_files/figure-latex/noBadge-1}

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_2.0_files/figure-latex/learnEvol-1}
